EBASE manuscript

### 2023-03-27 manu-draft.qmd
fix ebaseiotab using equatabs

```{r, results = 'hide'}
library(knitr)
library(here)
library(officer)
library(officedown)
library(flextable)

load(file = here('tabs/ebaseiotab.RData'))
load(file = here('tabs/apacmptab.RData'))
```

@@

<!-- edit -->

Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the NSE values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters. Changing the variance of the $a$ parameter had little effect on any of the comparisons. Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, although the precision was worse for the seven day optimization period (e.g., the $a$ parameter). Conversely, larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the seven day optimization period.

The results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions. The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-priorcomp. For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D). Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.

<!-- edit -->
### EBASE comparison with real observations

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for production (P), respiration (R), and gas exchange (D). The $a$ parameter is shown as an areal value for comparison with the metabolic results.](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day, seven day, and thirty day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.  The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

>>>>

```{r, results = 'hide'}
library(knitr)
library(here)
library(officer)
library(officedown)
library(flextable)
library(dplyr)

source(here('R/funcs.R'))

load(file = here('tabs/ebaseiotab.RData'))
load(file = here('tabs/apacmptab.RData'))
load(file = here('data/apasumdat.RData'))
```

@@

<!-- edit -->
```{r}
# percent negative nse not do
nseall <- apasumdat %>% 
  tidyr::unnest('ests') %>% 
  filter(var != 'DO_mod') %>% 
  na.omit %>% 
  pull(nse)
perneg <- (sum(nseall < 0) / length(nseall) * 100) %>% 
  round() %>% 
  as.character() %>% 
  paste0('%')

# gasex nse
prmall <- apasumdat %>% 
  tidyr::unnest('ests') %>% 
  filter(var != 'DO_mod') %>% 
  na.omit %>% 
  summarise(
    nse = round(median(nse), 2),
    .by = var
  ) %>% 
  tibble::deframe()

# optimization nse
optnse <- apasumdat %>% 
  tidyr::unnest('ests') %>% 
  na.omit() %>% 
  filter(var != 'DO_mod') %>% 
  summarise(
    nse = round(median(nse), 2), 
    .by = ndays
  ) %>% 
  tibble::deframe()

# best mods by opt
optmod <- metsum_fun(apasumdat, 'nse', parms = T) %>% 
  filter(rnkmetsum %in% 1) %>% 
  summarise(
    metsum = round(median(metsum), 2),
    .by  = 'ndays'
  ) %>% 
  tibble::deframe()
```

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.  The only models where the NSE values were less than zero for dissolved oxygen were those where $\rho$ of the $r$ parameter was low.  However, substantial variation in NSE values was observed among all other parameters.  Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`).  Conversely, the $a$ parameter was reproduced the most poorly, with a median value across all optimization periods and priors less than zero (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters decreased slightly as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).  The top three ranked models were identical for the seven and thirty day models and similar for models using the one day optimization period.  Median NSE values across parameters for the top model in each optimization period were also similar (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE < 0. Most parameters regardless of optimization period or priors had NSE < 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\rho$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.  Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.  For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\rho$ value for the $a$ prior distribution was low or high.  The size of the points indicate the inter-quartile range of NSE values for each subset.  As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values. Although most most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.  For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value.  A similar, but smaller effect is seen for the $\rho$ value of the prior for the $b$ parameter.  A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.  Conversely, model results are less senstive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.  This suggests that EBASE model sensitivity is most influenced by $\rho$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameters estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.  However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).  These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model.  Results using EBASE default priors (medium values for $\mu$, high values for $\rho$) showed that the $b$ parameter can be estimated with reasonable accuracy, although still with considerable variability.  Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).  The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\rho$.

### EBASE comparison with real observations

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay. Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for production (P), respiration (R), and gas exchange (D). The $a$ parameter is shown as an areal value for comparison with the metabolic results.  The $b$ parameter is also fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ and is not shown.](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combinations of prior distributions and optimization period. The means and standard deviations are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.  The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
